---
title: "20236 Time Series Analysis - Assignment 4"
author:
- Simone Arrigoni (1794692)
- Luca Badolato (3086040)
- Simone Valle (3088281)
subtitle: "Bocconi University"
date: April 4, 2020
output: 
  pdf_document
header-includes:
  \usepackage[utf8]{inputenc}
  \usepackage{setspace}
  \usepackage{algpseudocode}
  \usepackage{algorithm}
  \usepackage{bm}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{graphicx}
  \usepackage{subfig}
  \usepackage{booktabs, caption}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{listings}
  \usepackage{physics}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{color} %red, green, blue, yellow, cyan, magenta, black, white
  \definecolor{mygreen}{RGB}{28,172,0} % color values Red, Green, Blue
  \definecolor{mylilas}{RGB}{170,55,241}
  \DeclareMathOperator*{\E}{\mathbb{E}}
  \DeclareMathOperator*{\Ec}{\mathbb{E}_t}
---

```{r, include=FALSE}

# Load useful packages
library(utf8)
library(labeling)
library(rmarkdown)
library(httr)
library(knitr)
library(tseries)
library(tinytex)
library(scales)
library(dlm)
library(magrittr)
library(stringr)
library(depmixS4)
library(tidyverse)
library(ggthemes)
library(latex2exp)
library(kableExtra)
library(ggpubr)
library(reshape2)

# Settings
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      echo    = FALSE,
                      include = TRUE,
                      fig.pos = "H",
                      fig.align = "center")
```

\section{Exercise}

In this exercise we use the dataset \texttt{data\_a4.csv}, which contains monthly data for the period January 1997 to July 2019 including 10 years Italian government bond's interest rate, inflation represented by the Harmonised Index of Consumer Prices (HICP) and default ratings assigned by Moody's and Fitch in the investment grade range.
The structure of the dataset can be seen in Table \ref{tab:data}, where the first observations are listed.

```{r echo=FALSE}
# Import the dataset
data_a4 <- read.csv(url("https://raw.githubusercontent.com/LucaBadolato/TimeSeries-/master/data_a4.csv"), sep = ",", header=T)
data_a4$Time <- as.Date(data_a4$Time, tryFormats = c("%Y_%m_%d"), optional = FALSE)
```
```{r}
# How the dataset looks like
knitr::kable(data_a4[1:5,], "latex", booktabs=T,
             caption="Dataset structure\\label{tab:data}") %>%
  kable_styling(latex_options="hold_position")
```

\begin{enumerate}

\item %1

Let's consider the \textit{real} interest rate for the 10 years Italian government bond, i.e. the ratio between the \textit{nominal} interest rate and the Harmonised Index of Consumer Prices (HICP).
Figure \ref{fig:ITAbond_real} plots the time series for the \textit{real} interest rate.

```{r, fig.width=6, fig.height=2.5, fig.cap="\\label{fig:ITAbond_real}Real interest rate for the 10 years Italian government bond"}

# Time series plot
ggplot(data_a4, aes(x=Time, y=ITAbond_real)) +
  geom_line(color="steelblue") +
  theme_classic(base_size=9) +
  labs(y="Real interest rate", x="Date") +
  scale_x_date(date_breaks="2 years", date_minor_breaks="1 year",
               limits=as.Date(c('1997_01_01','2019_07_01'), tryFormats=c("%Y_%m_%d"),
                              optional=FALSE), labels=date_format("%m/%Y")) +
  scale_y_continuous(breaks = seq(0,6,0.5)) +
  theme(axis.text.x=element_text(angle=45, hjust=1),
        axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

The time series is clearly non-stationary and there seems to be several change points, the most clear of which being at the end of 2008, in correspondence to the beginning of the Great recession. In light of the presence of these structural breaks, a transformation of the series into a stationary one is also unlikely to be carried out. For these reasons, a Hidden Markov Model (HMM) could be a reasonable choice for modelling this time series.

Let's indeed model the series using a HMM. Specifically, in this exercise we use a homogeneous HMM, namely we assume that the hidden process $(S_{t})_{t\geq0}$ is a homogeneous Markov chain.

\textbf{Step 1: Model specification}

Let's assume:
\begin{itemize}
\item three states, representing respectively boom (low interest rates), recession (high interest rates) and a stable path;
\item Gaussian emission distributions, with state-dependent mean $\mu_i$ and variance $\sigma_i$ with $i=1,2,3$.
\end{itemize}

Moreover, no linear trend clearly shows up from the time series plot at Figure \ref{fig:ITAbond_real} suggesting that the data oscillates around a reference value that, as explained before, seem to depend on a latent state variable (or \textit{regime}).
Given the previous considerations, we shall specify for the interest rate a Gaussian HMM in its simplest form, i.e. a long-term state-dependent mean plus noise with state-dependent volatility.

\begin{align*}
\begin{cases}
Y_{t}=\mu_{1}+\varepsilon_{t}, \quad \varepsilon_{t}\overset{i.i.d.}{\sim}N(0,\sigma_{1}^2) \qquad \text{if} \; S_{t}=1 \\
Y_{t}=\mu_{2}+\varepsilon_{t}, \quad \varepsilon_{t}\overset{i.i.d.}{\sim}N(0,\sigma_{2}^2) \qquad \text{if} \; S_{t}=2 \\
Y_{t}=\mu_{3}+\varepsilon_{t}, \quad \varepsilon_{t}\overset{i.i.d.}{\sim}N(0,\sigma_{3}^2) \qquad \text{if} \; S_{t}=3
\end{cases}
\end{align*}

```{r}
# 1) Model specification
y <- as.numeric(data_a4$ITAbond_real)
nstates <- 3
model <- depmixS4::depmix(y ~ 1, data=data.frame(y), nstates=nstates)
```

\textbf{Step 2: Estimation of the unknown parameters}

We can now estimate the MLE of the unknown parameters of the model.
Table \ref{tab:HMM_MLE} presents the estimates the mean and standard deviation in the three states as well as their standard errors.
\medskip

\textcolor{red}{Quelli riportati nella tabella seguente (ottenuta su LaTeX inserendo manualmente i valori) sono i valori corretti. È possibile che nella tabella ancora seguente (ottenuta su R con "kable" salvando i valori nell'environment) le righe siano invertite. Non c'è problema, finché non assegniamo a ciascuno stato uno specifico market regime (e.g., S=1 è recessione), cosa che non abbiamo fatto.}
\begin{table}[H]
\centering
{\color{red}\begin{tabular}{ccc}
\toprule
State & $\hat{\mu}_{MLE}$ & $\hat{\sigma}_{MLE}$ \\
\midrule
S=1   & 1.29               & 0.351                 \\
      & (0.067)            & (0.033)               \\
S=2   & 3.169              & 0.737                 \\
      & (0.071)            & (0.049)               \\
S=3   & 2.019              & 0.207                 \\
      & (0.055)            & (0.033) \\ \bottomrule
\end{tabular}}
\end{table}

```{r include=FALSE}
# 2) Estimation of the unknown parameters
fmodel <- depmixS4::fit(model)
states_mu_sigma <- depmixS4::summary(fmodel)
states_mu_sigma
```
```{r}
# MLE
mu1hat <- paste0(round(fmodel@response[[1]][[1]]@parameters$coefficients,3))
mu2hat <- paste0(round(fmodel@response[[2]][[1]]@parameters$coefficients,3))
mu3hat <- paste0(round(fmodel@response[[3]][[1]]@parameters$coefficients,3))
sd1hat <- paste0(round(fmodel@response[[1]][[1]]@parameters$sd,3))
sd2hat <- paste0(round(fmodel@response[[2]][[1]]@parameters$sd,3))
sd3hat <- paste0(round(fmodel@response[[3]][[1]]@parameters$sd,3))
# se(MLE)
MLEse=depmixS4::standardError(fmodel)
se_mu1hat <- paste0('(', round(MLEse$se[13],3), ')')
se_mu2hat <- paste0('(', round(MLEse$se[15],3), ')')
se_mu3hat <- paste0('(', round(MLEse$se[17],3), ')')
se_sd1hat <- paste0('(', round(MLEse$se[14],3), ')')
se_sd2hat <- paste0('(', round(MLEse$se[16],3), ')')
se_sd3hat <- paste0('(', round(MLEse$se[18],3), ')')
```
```{r}
# Build a summary table
MLEsum <- data.frame(state=c(rep("S=1",2), rep("S=2",2), rep("S=3",2)),
                     mu=c(mu1hat, se_mu1hat, mu2hat, se_mu2hat, mu3hat, se_mu3hat),
                     sd=c(sd1hat, se_sd1hat, sd2hat, se_sd2hat, sd3hat, se_sd3hat))
knitr::kable(MLEsum, "latex", booktabs=T, align="c",
             col.names = c("State", "$\\hat{\\mu}_{MLE}$", "$\\hat{\\sigma}_{MLE}$"),
             escape=F, caption="Maximum Likelihood estimates of the state-dependent
             mean and standard deviation\\label{tab:HMM_MLE}") %>%
  column_spec(1:3, width="5em") %>%
  collapse_rows(columns=1, latex_hline="none") %>%
  footnote(general="Standard errors in parentheses",
           general_title="Note:",
           footnote_as_chunk=T, escape=F) %>%
  kable_styling(latex_options="hold_position")
```

\textbf{Step 3: Decoding}

Finally, we find the optimal state variable path given the observed real interest rates, that is, we proceed with a global decoding procedure based on the \textit{Viterbi algorithm}.
Viterbi states are plotted in Figure \ref{fig:model_fit_ost}.

```{r}
# Get the estimated state for each timestep 
estStates <- posterior(fmodel)
```

```{r, fig.width=6, fig.height=3, fig.cap="\\label{fig:model_fit_ost}Optimal state trajectory"}

# Plot the optimal path, that is stored in the first column of estStates
results_df <- data.frame(time_index=data_a4$Time,
                         interest_rate=data_a4$ITAbond_real,
                         state=estStates[1]) %>% 
  gather("variable", "value", -time_index)

my_breaks <- function(x) { if (max(x)<4) seq(0,3,1) else seq(0,6,1) }
lab_names <- c(
  'interest_rate'="Real interest rate", 
  'state'="Viterbi state")

plotobj <- ggplot(results_df, aes(time_index, value)) +
  geom_line(color="black") +
  facet_wrap(variable~., scales="free", ncol=1,
             strip.position="left",
             labeller = as_labeller(lab_names)) +
  labs(x="Date") +
  scale_y_continuous(breaks=my_breaks) +
  scale_x_date(date_breaks="3 years",
               date_minor_breaks="1 year",
               limits=as.Date(c('1997_01_01','2019_07_01'),
                              tryFormats=c("%Y_%m_%d"),
                              optional=FALSE),
               labels=date_format("%Y-%m")) +
  theme_classic(base_size=9) +
  theme(axis.title.x=element_text(family="Times", margin=margin(5,0,0,0)),
        axis.title.y=element_blank(),
        strip.text.x=element_text(family="Times", face="bold",
                                  margin=margin(10,0,0,0)),
        strip.background=element_rect(colour="black", fill=NA),
        strip.placement="outside")

plot(plotobj)
```

The interest rates series with HMM estimated state-dependent means are shown in Figure \ref{fig:model_fit_rsm}.

```{r, fig.width=6, fig.height=2.5, fig.cap="\\label{fig:model_fit_rsm}Regime switching mean"}

# Chart with state-dependent mean
estMeans=matrix(0, length(data_a4[,1]),3)

for (i in 1:3) {
  for (j in 1:length(data_a4[,1])) {
    if (estStates[j,1]==i){
      estMeans[j,i]<-fmodel@response[[i]][[1]]@parameters$coefficients
   }else{
      estMeans[j,i]<-0
   }
  }
}

switching_mean <- estMeans[,1]+estMeans[,2]+estMeans[,3]

ggplot(data.frame(switching_mean, data_a4[1],data_a4[6])) +
  geom_line(aes(x=Time, y=ITAbond_real)) +
  geom_point(aes(x=Time, y=switching_mean), col="blue", size=.05) +
  theme_classic(base_size=9) +
  theme(plot.title=element_text(hjust=0.5)) +
  labs(y="Real interest rate", x="Date") +
  scale_x_date(date_breaks="2 years",
               date_minor_breaks="1 year",
               limits=as.Date(c('1997_01_01','2019_07_01'),
                              tryFormats=c("%Y_%m_%d"),
                              optional=FALSE),
               labels=date_format("%Y-%m")) +
  scale_y_continuous(breaks = seq(0,6,0.5)) +
  theme(plot.title=element_text(size=10, face="bold", margin=margin(0,0,10,0)),
        axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5))) +
  ggtitle("State dependent mean")
```

\textcolor{red}{An alternative (more complete) way of plotting the state-dependent means and standard deviations in shown in Figure \ref{fig:model_fit_rsm_2}.}

```{r, fig.width=6, fig.height=2.5, fig.cap="\\label{fig:model_fit_rsm_2}State dependent means and standard deviations"}

# Chart with state-dependent mean and standard deviations
info_df <- data.frame(state=1:nstates,
                      mu=states_mu_sigma[,1],
                      sigma=states_mu_sigma[,2])
df_to_plot <- estStates %>%
  left_join(info_df)
df_to_plot %<>%
  mutate(xtime=data_a4$Time, yvalue=data_a4$ITAbond_real)

ggplot(df_to_plot, aes(x=xtime, y=yvalue)) +
  geom_line() +
  geom_point(aes(x=xtime, y=mu), col="blue", size=.05) +
  geom_ribbon(aes(ymin=mu-2*sigma, ymax=mu+2*sigma), alpha=.1) +
  theme_classic(base_size=9) +
  theme(plot.title=element_text(hjust=0.5)) +
  labs(y="Real interest rate", x="Date") +
  scale_x_date(date_breaks="2 years",
               date_minor_breaks="1 year",
               limits=as.Date(c('1997_01_01','2019_07_01'),
                              tryFormats=c("%Y_%m_%d"),
                              optional=FALSE),
               labels=date_format("%Y-%m")) +
  scale_y_continuous(breaks = seq(0,6,0.5)) +
  theme(plot.title=element_text(size=10, face="bold", margin=margin(0,0,10,0)),
        axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5))) +
  ggtitle("State dependent mean")
```

\item %2
We now use a non-homogeneous HMM, namely we allow the transition matrix to depend on covariates, in order to \textit{predict} possible change points in our time series.

```{r}
# 1) Model specification for non-homogeneous HMM
y_nh <- as.numeric(data_a4$ITAbond_real)
model_nh <- depmixS4::depmix(y_nh ~ 1, data=data.frame(y_nh), nstates=3, transition=~ y_nh)
```

We proceed exactly as above and first estimate the MLE of the unknown parameters of the model.
Table \ref{tab:HMM_MLE_nh} presents the estimates the mean and standard deviation in the three states as well as their standard errors.

```{r include=FALSE}
# 2) Model estimation for non-homogeneous HMM
fmodel_nh <- depmixS4::fit(model_nh)
depmixS4::summary(fmodel_nh)
```
```{r}
# MLE
mu1hat_nh <- paste0(round(fmodel_nh@response[[1]][[1]]@parameters$coefficients,3))
mu2hat_nh <- paste0(round(fmodel_nh@response[[2]][[1]]@parameters$coefficients,3))
mu3hat_nh <- paste0(round(fmodel_nh@response[[3]][[1]]@parameters$coefficients,3))
sd1hat_nh <- paste0(round(fmodel_nh@response[[1]][[1]]@parameters$sd,3))
sd2hat_nh <- paste0(round(fmodel_nh@response[[2]][[1]]@parameters$sd,3))
sd3hat_nh <- paste0(round(fmodel_nh@response[[3]][[1]]@parameters$sd,3))
# se(MLE)
MLEse_nh=depmixS4::standardError(fmodel_nh)
se_mu1hat_nh <- paste0('(', round(MLEse_nh$se[22],3), ')')
se_mu2hat_nh <- paste0('(', round(MLEse_nh$se[24],3), ')')
se_mu3hat_nh <- paste0('(', round(MLEse_nh$se[26],3), ')')
se_sd1hat_nh <- paste0('(', round(MLEse_nh$se[23],3), ')')
se_sd2hat_nh <- paste0('(', round(MLEse_nh$se[25],3), ')')
se_sd3hat_nh <- paste0('(', round(MLEse_nh$se[27],3), ')')
```
```{r}
# Summary table creation
MLEsum <- data.frame(state=c(rep("S=1",2), rep("S=2",2), rep("S=3",2)),
                     mu=c(mu1hat_nh, se_mu1hat_nh, mu2hat_nh, se_mu2hat_nh,
                          mu3hat_nh, se_mu3hat_nh),
                     sd=c(sd1hat_nh, se_sd1hat_nh, sd2hat_nh, se_sd2hat_nh,
                          sd3hat_nh, se_sd3hat_nh))
knitr::kable(MLEsum, "latex", booktabs=T, align="c",
             col.names = c("State", "$\\hat{\\mu}_{MLE}$", "$\\hat{\\sigma}_{MLE}$"),
             escape=F, caption="Maximum Likelihood estimates of the state-dependent
             mean and standard deviation\\label{tab:HMM_MLE_nh}") %>%
  column_spec(1:3, width="5em") %>%
  collapse_rows(columns=1, latex_hline="none") %>%
  footnote(general="Standard errors in parentheses",
           general_title="Note:",
           footnote_as_chunk=T, escape=F) %>%
  kable_styling(latex_options="hold_position")
```

Then, in order to find the optimal state variable path, we perform the global decoding procedure based on the \textit{Viterbi algorithm}.
Viterbi states and state-dependent means in the non-homogeneous HMM case are plotted respectively in Figure \ref{fig:model_fit_ost_nh} and Figure \ref{fig:model_fit_rsm_nh}.

```{r}
# Get the estimated state for each timestep 
estStates_nh <- posterior(fmodel_nh)
```

```{r, fig.width=6.5, fig.height=3, fig.cap="\\label{fig:model_fit_ost_nh}Optimal state trajectory (non-homogeneous HMM)"}

# Plot the optimal path, that is stored in the first column of estStates
results_df_nh <- data.frame(time_index=data_a4$Time,
                         interest_rate=data_a4$ITAbond_real,
                         state=estStates_nh[1]) %>% 
  gather("variable", "value", -time_index)

my_breaks <- function(x) { if (max(x)<4) seq(0,3,1) else seq(0,6,1) }
lab_names <- c(
  'interest_rate'="Real interest rate", 
  'state'="Viterbi state")

plotobj_nh <- ggplot(results_df_nh, aes(time_index, value)) +
  geom_line(color="black") +
  facet_wrap(variable~., scales="free", ncol=1,
             strip.position="left",
             labeller = as_labeller(lab_names)) +
  labs(x="Date") +
  scale_y_continuous(breaks=my_breaks) +
  scale_x_date(date_breaks="3 years",
               date_minor_breaks="1 year",
               limits=as.Date(c('1997_01_01','2019_07_01'),
                              tryFormats=c("%Y_%m_%d"),
                              optional=FALSE),
               labels=date_format("%Y-%m")) +
  theme_classic(base_size=9) +
  theme(axis.title.x=element_text(family="Times", margin=margin(5,0,0,0)),
        axis.title.y=element_blank(),
        strip.text.x=element_text(family="Times", face="bold",
                                  margin=margin(10,0,0,0)),
        strip.background=element_rect(colour="black", fill=NA),
        strip.placement="outside")

plot(plotobj_nh)
```

```{r, fig.width=6, fig.height=2.5, fig.cap="\\label{fig:model_fit_rsm_nh}Regime switching mean (non-homogeneous HMM)"}

# Chart with state-dependent mean
estMeans_nh=matrix(0, length(data_a4[,1]),3)

for (i in 1:3) {
  for (j in 1:length(data_a4[,1])) {
    if (estStates_nh[j,1]==i){
      estMeans_nh[j,i]<-fmodel_nh@response[[i]][[1]]@parameters$coefficients
   }else{
      estMeans_nh[j,i]<-0
   }
  }
}

switching_mean_nh <- estMeans_nh[,1]+estMeans_nh[,2]+estMeans_nh[,3]

ggplot(data.frame(switching_mean_nh, data_a4[1],data_a4[6])) +
  geom_line(aes(x=Time, y=ITAbond_real)) +
  geom_point(aes(x=Time, y=switching_mean_nh), col="blue", size=.05) +
  theme_classic(base_size=9) +
  theme(plot.title=element_text(hjust=0.5)) +
  labs(y="Real interest rate", x="Date") +
  scale_x_date(date_breaks="3 years",
               date_minor_breaks="1 year",
               limits=as.Date(c('1997_01_01','2019_07_01'),
                              tryFormats=c("%Y_%m_%d"),
                              optional=FALSE),
               labels=date_format("%Y-%m")) +
  scale_y_continuous(breaks = seq(0,6,1)) +
  theme(plot.title=element_text(size=10, face="bold", margin=margin(0,0,10,0)),
        axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5))) +
  ggtitle("State dependent mean")
```
\end{enumerate}

\section{Final project}
\subsection{Introduction}

In this exercise we use the dataset \texttt{USPollution}, which contains monthly and daily data about air pollution in the United States for each country. The dataset is provided by the United States Environmental Protection Agency (EPA), an independent executive agency of the United States federal government, in charge of maintaining and enforcing national standards under a variety of environmental laws. Particularly, 4 key indicators of air pollution are considered: 

 \begin{itemize}
\item Ground level ozone (\texttt{O3}): bad ozone (there is also a good one naturally present in the Earth's upper atomsphere, which protect us from the sun's harmful ultraviolet rays) forms near the ground when pollutants (emitted by sources such as cars, power plants, industrial boilers, refineries, and chemical plants) react chemically in sunlight. Since it is more likely to form during warmer months, we expect to see a seasonal component in the series. 
\item Nitrogen dioxide (\texttt{NO2}): It is produced as a result of road traffic and other fossil fuel combustion processes. Furthermore, its presencce in air contributes to the formation and modification of other air pollutants, such as ozone and particle pollution. This last one consists of a mixture of solids and liquid droplets, some of them so small that can get into the lungs, causing serious health problems. 
\item Carbon monoxide (\text{CO2}):  It forms when the carbon in flues does not completely burn. The principal causes are given by vehicle exhaust (it contributes roughly 75 percent of all carbon monoxide emissions nationwide, up to 95% in cities) and fuel combustion in industrial processes and natural sources such as wildfires. Cold temperatures make combustion less complete and cause inversions that trap the CO2 close to the ground, therefore we expect also in this case a seasonal behavior. 
\item Sulfur dioxide (\texttt{SO2}): It is produced when sulfur-containing fuels such as coal and oil are burned. Tha major cause of this pullutant are large industrial complexes such as power plants, refineries and industrial boilers. 
\end{itemize}

Hence, the four variables of interest are interconnected and are the building blocks of a more general phenomenon, air pollution. Furthermore, they are related by a common latent process that can be summarized by several components: industrialization, car traffic and regulation. 

In our analysis we will focus on the four times series related to Alaska, recorded daily from *anno di inizio* to *anno fine*. The main advantage is given by study the evolution of air pollution in the same country over the years. This allows to control for fixed effects and to better study the latent process of interest that relate the four pollution components. 

