---
title: "20236 Time Series Analysis - Assignment 5"
author:
- Simone Arrigoni (1794692)
- Luca Badolato (3086040)
- Simone Valle (3088281)
subtitle: "Bocconi University"
date: April 15, 2020
output: pdf_document
header-includes:
  \usepackage[utf8]{inputenc}
  \usepackage{setspace}
  \usepackage{algpseudocode}
  \usepackage{algorithm}
  \usepackage{bm}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{graphicx}
  \usepackage{subfig}
  \usepackage{booktabs, caption}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{listings}
  \usepackage{physics}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{color} %red, green, blue, yellow, cyan, magenta, black, white
  \definecolor{mygreen}{RGB}{28,172,0} % color values Red, Green, Blue
  \definecolor{mylilas}{RGB}{170,55,241}
  \DeclareMathOperator*{\E}{\mathbb{E}}
  \DeclareMathOperator*{\Ec}{\mathbb{E}_t}
---

```{r, include=FALSE}

# Load useful packages
library(utf8)
library(labeling)
library(rmarkdown)
library(httr)
library(knitr)
library(tseries)
library(tinytex)
library(scales)
library(magrittr)
library(stringr)
library(tidyverse)
library(ggplot2)
library(ggfortify)
library(ggpubr)
library(ggthemes)
library(latex2exp)
library(kableExtra)
library(reshape2)
library(depmixS4)
library(dlm)

# Settings
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      echo    = FALSE,
                      include = TRUE,
                      fig.pos = "H",
                      fig.align = "center",
                      out.width = '0.6')
```

\section{Exercise}

In this exercise we consider the \textbf{Nile} data, available in R. This provides the measurements of the annual flow of the river Nile at Ashwan, for the period 1871-1970.

We first plot the data in Figure \ref{fig:Nile_plot}.

The series is clearly non-stationary and presents a break point around 1998 (red vertical line).

```{r}
# Plot the Series "Nile"
# ts.plot(Nile, gpars = list(xlab="Year", ylab="10^8 m^3"))

years_Nile<-seq(as.Date("1871/01/01"),as.Date("1970/01/01"), "years")

df_nile<-data.frame(years_Nile,Nile)
   
ggplot(df_nile, aes(x=years_Nile, y=Nile)) +
  geom_line() +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "red")+
  theme_classic(base_size=9) +
  theme(plot.title=element_text(hjust=0.5)) +
  labs(y="10^8 m^3", x="Year") +
  scale_x_date(date_breaks="20 years",
               date_minor_breaks="10 year",
               limits=as.Date(c('1871_01_01','1970_01_01'),
                              tryFormats=c("%Y_%m_%d"),
                              optional=FALSE),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(0,1500,300), limits = c(300,1500)) +
  theme(axis.text.x=element_text(hjust=1),
        plot.title=element_text(size=10, face="bold", margin=margin(0,0,10,0)),
        axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

\subsection{Question 1}

The random walk plus noise (or local level) model is a time-invariant DLM with univariate Gaussian observation and state process, i.e. $\underset{1 \times 1}{Y_t}$ and $\underset{1 \times 1}{\theta_t}$, and with constant system matrices $F=1, G=1$. The variances of these processes are constant and set at $V=\sigma^2_v=15100$ $W=\sigma^2_w=1470$. The initial distribution as well is assumed to be Gaussian with parameters $m_0=1000$ and $C_0=1000$.

```{r}
# Polinomial specification of the RW + noise

rw_Nile <- dlmModPoly(order=1,dV=15100, dW=1470, m0=1000, C0=1000)
filt_Nile <- dlmFilter(Nile,rw_Nile)

# Remove the first item (m_0) in the vector of filtered states
filt_est=dropFirst(filt_Nile$m)

#compute the vector of state variances from the Singular Variance Decomposition matrices
list_c <- dlmSvd2var(filt_Nile$U.C, filt_Nile$D.C)

# Compute standard deviations
vol_C <- sqrt(unlist(list_c))


```
