---
title: "20236 Time Series Analysis - Assignment 6"
author:
- Simone Arrigoni (1794692)
- Luca Badolato (3086040)
- Simone Valle (3088281)
subtitle: "Bocconi University"
date: May 4, 2020
output: pdf_document
header-includes:
  \usepackage[utf8]{inputenc}
  \usepackage{setspace}
  \usepackage{apacite}
  \usepackage{natbib}
  \usepackage{algpseudocode}
  \usepackage{algorithm}
  \usepackage{bm}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{graphicx}
  \usepackage{subfig}
  \usepackage{booktabs, caption}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{listings}
  \usepackage{physics}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{color} %red, green, blue, yellow, cyan, magenta, black, white
  \definecolor{mygreen}{RGB}{28,172,0} % color values Red, Green, Blue
  \definecolor{mylilas}{RGB}{170,55,241}
  \definecolor{codegray}{RGB}{0.5,0.5,0.5}
  \definecolor{orange}{RGB}{255,165,0}
  \DeclareMathOperator*{\E}{\mathbb{E}}
  \DeclareMathOperator*{\Ec}{\mathbb{E}_t}
  \setlength\parindent{0pt}
  \newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
  \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
---

```{r, include=FALSE}
# Load useful packages
library(utf8)
library(labeling)
library(rmarkdown)
library(httr)
library(knitr)
library(tseries)
library(tinytex)
library(scales)
library(dlm)
library(magrittr)
library(stringr)
library(depmixS4)
library(tidyverse)
library(ggthemes)
library(latex2exp)
library(kableExtra)
library(ggpubr)
library(reshape2)
library(dlm)
# Settings
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      echo    = FALSE,
                      include = TRUE,
                      fig.pos = "H",
                      fig.align = "center",
                      out.width='60%')
```

\section{Exercise}

In this exercise we implement in R a DLM for regression. In particular, we consider the **capital asset pricing model (CAPM)**, a popular model in financial applications.

This model assumes that the *excess returns* $Y_t$ of a financial asset depend linearly on the *market excess returns* $x_t$, that is:
\begin{eqnarray*}
Y_t = \alpha + \beta x_t + v_t \qquad v_t \overset{i.i.d.}\sim N(0, \sigma^2)
\end{eqnarray*}
where
\begin{itemize}
\item $Y_t= r_t - r_t^{(f)}$, with $r_t$ and $r_t^{(f)}$ being the returns at time $t$ of the asset of interest and of a risk-free asset, respectively;  
\item $x_t = r_t^{(M)}- r_t^{(f)}$, with $r_t^{(M)}$ being the return at time $t$ of the market.
\end{itemize}

The dataset we use includes the monthly returns from January 1978 to December 1987 of four stocks (Mobil, IBM, Weyer, Citicorp), and the 30-days Treasury Bill as a proxy for the risk-free asset, and of the value-weighted average returs for all the stocks listed at the New York and American Stock Exchanges, representing the overall market returns.

```{r, include=FALSE}
CAPM = read.table("dati-ch3-CAPM.txt", header=TRUE)
is.ts(CAPM[,"WEYER"])
CAPM = ts(CAPM, frequency=12, start=c(1978,1))
```

The time series of the stocks are plotted in Figure \ref{fig:stock_plots}

```{r, fig.cap="\\label{fig:stock_plots} Stock plots"}
# Plot
ggplot(CAPM) +
  geom_line(color="black", size=0.4) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y="Return") +
  scale_x_date(date_breaks="1 year", limits=as.Date(c('1978/01/01','1987/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous() +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

We will focus on the Weyer stock.

Let us now write the CAPM as a DLM with static coefficients.
This means writing the linear regression in the following DLM form:

\begin{eqnarray*} 
y_t &=& \alpha_t + \beta_t x_t + v_t , \quad v_t \overset{i.i.d.}\sim  N(0 , \sigma^2_v) \\
\theta_t &=& G \theta_{t-1} + w_t , \quad w_t \overset{i.i.d.}\sim N_2 (0, W)
\end{eqnarray*}
where
\begin{eqnarray*}
\theta_t=[\alpha_t, \beta_t]', \qquad  \theta_0 \sim N_2(m_0, C_0), \qquad \theta_0 \independent (v_t) \independent (w_t)
\end{eqnarray*}

Let us use the simple specification $G=I_2$, $W=diag(\sigma^2_{w_1}, \sigma^2_{w_2})$, that is, $(\alpha_t)$ and $(\beta_t)$ are independent random walks.
For static regression coefficients, we set $\sigma^2_{w_1}=0$, $\sigma^2_{w_2}=0$.

```{r, include=FALSE}
# Returns
y = CAPM[, "WEYER"] - CAPM[, "RKFREE"]  # IMB excess returns
x = CAPM[, "MARKET"]- CAPM[, "RKFREE"]  # market excess returns
```

```{r}
# 
lmCAPM = lm(y ~ x)
lmCAPM$coef
# DLM
buildCAPM <- function(param){dlmModReg(x, addInt=TRUE, dV=param, dW=c(0,0))}
# MLE
fit=dlmMLE(y, parm=1, buildCAPM)
```



