---
title: "20236 Time Series Analysis - Assignment 5"
author:
- Simone Arrigoni (1794692)
- Luca Badolato (3086040)
- Simone Valle (3088281)
subtitle: "Bocconi University"
date: April 15, 2020
output: pdf_document
header-includes:
  \usepackage[utf8]{inputenc}
  \usepackage{setspace}
  \usepackage{algpseudocode}
  \usepackage{algorithm}
  \usepackage{bm}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{graphicx}
  \usepackage{subfig}
  \usepackage{booktabs, caption}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{listings}
  \usepackage{physics}
  \usepackage{float}
  \floatplacement{figure}{H}
  \usepackage{color} %red, green, blue, yellow, cyan, magenta, black, white
  \definecolor{mygreen}{RGB}{28,172,0} % color values Red, Green, Blue
  \definecolor{mylilas}{RGB}{170,55,241}
---

```{r, include=FALSE}

# Load useful packages
library(utf8)
library(labeling)
library(rmarkdown)
library(httr)
library(knitr)
library(tseries)
library(tinytex)
library(scales)
library(magrittr)
library(stringr)
library(tidyverse)
library(ggplot2)
library(ggfortify)
library(ggpubr)
library(ggthemes)
library(latex2exp)
library(kableExtra)
library(reshape2)
library(depmixS4)
library(dlm)

# Settings
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      echo    = FALSE,
                      include = TRUE,
                      fig.pos = "H",
                      fig.align = "center")
```

\section{Exercise}

In this exercise we consider the \textbf{Nile} time series, available in R, which contains the measurements of the annual flow of the river Nile at Ashwan, for the period 1871-1970.

We first plot the data in Figure \ref{fig:Nile_plot}.
\bigskip

```{r, fig.width=5, fig.height=2, fig.cap="\\label{fig:Nile_plot}Annual flow of the river Nile at Ashwan"}
# Df
years_Nile<-seq(as.Date("1871/01/01"),as.Date("1970/01/01"), "years")
df_Nile<-data.frame(years_Nile,Nile)

# Plot
ggplot(df_Nile, aes(x=years_Nile, y=Nile)) +
  geom_line(color="black", size=0.4) +
  geom_vline(xintercept=as.Date('1899/01/01'), color="black", size=0.3)+
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y=expression('Annual flow (10'^"8"*' m'^"3"*')')) +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(500,1500,300), limits = c(400,1500)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

The series is clearly non-stationary and presents a change point around 1998 (black vertical line). For this reason, studying it fitting a local level model seems a good idea.

\subsection{Question 1}

Let's indeed model the series with a random walk plus noise (or local level) model. This is a time-invariant DLM with univariate Gaussian observation and state process, i.e. $\underset{1 \times 1}{Y_t}$ and $\underset{1 \times 1}{\theta_t}$, as well as with constant system matrices $\underset{1 \times 1}F=1$ and $\underset{1 \times 1}G=1$:
\begin{eqnarray*} 
& Y_t = \theta_t + v_t \quad            & v_t \overset{i.i.d.}\sim N(0, V) \\
& \theta_t = \theta_{t-1} + w_t \quad   & v_t \overset{i.i.d.}\sim N(0, W)
\end{eqnarray*}
where the variances of both processes are constant and set at $V=\sigma^2_v=15100$ and $W=\sigma^2_w=1470$ respectively.

The prior distribution is also assumed to be Gaussian with parameters $m_0=1000$ and $C_0=1000$, that is:
\begin{align*} 
\theta_0 \sim N(1000, 1000)
\end{align*}

Our first goal is to compute the \textit{filtering states estimates} $m_t =\mathbb{E}\;(\theta_t \mid y_{1:t})$ obtained by implementing the Kalman filter. In particular, we recall that the Kalman filter procedure, given the \textit{filtering distribution} in the previous iteration, $\theta_{t-1} \mid y_{1:t-1} \sim N(m_{t-1},C_{t-1})$, follows three steps:

\begin{itemize}
\item State forecasting step: 
$\theta_t \mid y_{1:t-1} \sim N(a_t, R_t), \quad \mbox{state prediction}$
\item Observation forecasting step:
$Y_t \mid y_{1:t-1} \sim N(f_t, Q_t), \quad \mbox{observation prediction}$
\item Filtering step: once $y_t$ becomes available
$\theta_t \mid y_{1:t} \sim N(m_t, C_t), \quad \mbox{filtering distribution}.$
\end{itemize}

```{r}
# Polinomial specification of the RW + noise
rw_Nile <- dlmModPoly(order=1,dV=15100, dW=1470, m0=1000, C0=1000)

# Kalman filter
filt_Nile <- dlmFilter(Nile,rw_Nile)

# Remove the first item (m_0) in the vector of filtered states
filt_est=dropFirst(filt_Nile$m)

# Compute the vector of state variances from the Singular Variance Decomposition matrices
list_c <- dlmSvd2var(filt_Nile$U.C, filt_Nile$D.C)

# Compute standard deviations
vol_C <- sqrt(unlist(list_c))

```
\medskip

We report in Figure \ref{fig:filter_est} the \textit{filtering states estimates} $(m_1, \ldots, m_T)$ (red line) together with the observed series (black line).
Note that the initial guess on the location parameter of the state distribution, $m_0$, has to be removed from the series of states estimates and is, therefore, not included in the plot. 
\bigskip

```{r, fig.width=7, fig.height=2, fig.cap="\\label{fig:filter_est} Filtering states estimates"}
# Df
df_Nile_filt <-data.frame(years_Nile, Nile, filt_est)

# Plot
colors <- c("Observed data" = "black", "Filtering estimates" = "red")
ggplot(df_Nile_filt, aes(x=years_Nile)) +
  geom_line(aes(y=Nile, color="Observed data"), size=0.4) +
  geom_line(aes(y=filt_est, color = "Filtering estimates"), size=0.4) +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y="State level", color = "") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(500,1500,300), limits = c(400,1500)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))+
        scale_color_manual(values = colors)+
        theme(legend.justification = c("left", "top"),
              legend.box.just = "left",legend.margin = margin(2, 2, 2, 2))
```

Then, we compute the \textit{filtering variances} $(C_1, \ldots, C_T)$.
In principle, they could be computed through the recursive formulas of Kalman filter. However, due to potential numerical issues in the application of such recursive approach, the computation of state variances relies on the Singular Value Decomposition procedure.

In Figure \ref{fig:filter_sd} we report the filtered standard deviations $\sqrt{C_{t}} = V(\theta_t \mid y_{1:t-1})^{\frac{1}{2}}$. 
\bigskip

```{r}
vol_C_plot <- vol_C[2:101]
```

```{r, fig.width=5, fig.height=2, fig.cap="\\label{fig:filter_sd} Standard deviations of states estimates"}

df_Nile_sd <-data.frame(years_Nile, vol_C_plot)

 ggplot(df_Nile_sd, aes(x=years_Nile))+
  geom_line(aes(y=vol_C_plot), color = "darkgreen", size=0.4) +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y="Standard deviation") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(40,70,10), limits = c(40,70)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

Interestingly, the filtered standard deviations are increasing at a decreasing pace until converging to a limit value equal to `r round(vol_C_plot[7], 2)` after six iterations.

Finally, we compute the $(1-\alpha)$ Bayesian credible intervals for the filtering state estimates, which are given by:
\begin{align*}
\bigg[ m_t - Z_{1 - \frac{\alpha}{2}} \sqrt{C_{t}}, \quad m_t - Z_{1 - \frac{\alpha}{2}}\sqrt{C_{t}} \bigg]
\end{align*}
with $\alpha=0.05$.

We plot in Figure \ref{fig:filter_est_ci} the observations with their filtering state estimates and their 95% level credible intervals.
\bigskip

```{r, fig.width=7, fig.height=2, fig.cap="\\label{fig:filter_est_ci} Filtering estimates with credible intervals"}

upper <- filt_est + 1.96*vol_C_plot 
lower <- filt_est - 1.96*vol_C_plot 

df_Nile_ci <-data.frame(years_Nile, Nile, filt_est, upper, lower)
   
colors <- c("Observed data" = "black", "Filtering estimates" = "red")

 ggplot(df_Nile_sd, aes(x=years_Nile))+
  geom_line(aes(y=Nile, color = "Observed data"), size=0.3) +
  geom_line(aes(y=filt_est, color = "Filtering estimates with C.I."), size=0.5) +
  geom_line(aes(y=upper), color = "red", linetype = "dashed", size=0.1) +
  geom_line(aes(y=lower), color = "red", linetype = "dashed", size=0.1) +
  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=.1, fill="red") +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y=expression('Annual flow (10'^"8"*' m'^"3"*')'), color = "") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(500,1500,300), limits = c(400,1500)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))+
        theme(legend.justification = c("left", "top"),
              legend.box.just = "left",legend.margin = margin(2, 2, 2, 2))
```

\subsection{Question 2}

We now plot in Figure \ref{fig:forecast_ci} the Nile time series alongside with the one-step-ahead forecasts for the observation process (blue line), i.e. $\hat{y}_{t}=\mathbb{E}\;({Y_{t} \mid y_{1:t-1})}=f_t$, where $f_t$ is the location parameter of the observation's forecast distribution. The credible intervals is also showed, where the credible intervals at the $(1-\alpha)$ level are given by:

\begin{align*}
\bigg[ f_t - Z_{1 - \frac{\alpha}{2}} \sqrt{Q_{t}}, \quad f_t - Z_{1 - \frac{\alpha}{2}}\sqrt{Q_{t}} \bigg]
\end{align*}

with $\alpha=0.05$.

The variances $Q_t's$ are just a byproduct of the procedure outlined in Question 1. Indeed, they are obtained as:

\begin{align*}
Q_t = F_t \, R_t \, F_{t}'+ V_t = C_{t-1}+\sigma_{v}^2+\sigma_{w}^2
\end{align*}

with $F_t=1$.
\bigskip

```{r}
# Create the vector of means for the observation process.
# Notice that here we do not have to remove the first element of the series
mean_forecast<-filt_Nile$f

# Obtain the variance Q
filt_Q <- unlist(list_c) + as.vector(rw_Nile$V) + as.vector(rw_Nile$W)
filt_Q <- filt_Q[-1]

# Create upper and lower bounds of the credible interval
upper_for <- mean_forecast + 1.96*sqrt(filt_Q) 
lower_for <- mean_forecast - 1.96*sqrt(filt_Q) 
```

```{r, fig.width=7, fig.height=2, fig.cap="\\label{fig:forecast_ci} One-step ahead forecasts with credible intervals"}
# Df
df_Nile_for <-data.frame(years_Nile, Nile, mean_forecast, upper_for, lower_for)

colors <- c("Observed data" = "black", "Forecast with C.I." = "blue")

# Plot
ggplot(df_Nile_for, aes(x=years_Nile)) +
  geom_line(aes(y=Nile, color = "Observed data"), size=0.3) +
  geom_line(aes(y=mean_forecast, color = "Forecast with C.I."), size=0.5) +
  geom_line(aes(y=upper_for), color = "blue", linetype = "dashed", size=0.1) +
  geom_line(aes(y=lower_for), color = "blue", linetype = "dashed", size=0.1) +
  geom_ribbon(aes(ymin=lower_for, ymax=upper_for), alpha=.1, fill="blue") +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y=expression('Annual flow (10'^"8"*' m'^"3"*')'), color = "") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(500,1500,300), limits = c(400,1500)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))+
      theme(legend.justification = c("left", "top"),
              legend.box.just = "left",legend.margin = margin(2, 2, 2, 2))
```

\subsection{Question 3}

The behavior of the process $(Y_t)$ is highly influenced by the ratio between the variances of the state and the observations error, $r=\frac{W}{V}=\frac{\sigma^2_w}{\sigma^2_v}$, that is the so-called \textit{signal-to-noise ratio}.

Let us now define the \textit{adaptive coefficient}, $K_t$, as

\begin{align*}
  K_t = R_t F'_t Q_t^{-1}
\end{align*}

which in our simplified case is just

\begin{align*}
  K_t = \frac{R_t}{R_t + \sigma_{v}^{2}} = \frac{C_{t-1}+\sigma_{w}^{2}}{C_{t-1}+\sigma_{w}^{2}+\sigma_{v}^{2}}
\end{align*}

The adaptive coefficient, that clearly depends on the signal-to-noise ratio, is the weight associated to the error correction in the equation of the posterior mean obtained at the filtering step of the Kalman filter, i.e.

\begin{align*}
  m_t = a_t + K_t (y_t - a_t)
\end{align*}

where $(y_t - a_t)$ is, in fact, the error correction.

Therefore, for any given $C_0$, if the signal-to-noise ratio is small, $K_t$ is small and as a consequence $y_t$ receives little weight as error correction. On the contrary, if the ratio is large, $y_t$ receives large weight and the one-step-ahead forecast is close to the most recent data point.

We repeat the same analysis of Questions 1 and 2 choosing two different values for the abovementioned variances in order to find evidence that this is actually the case.
\bigskip

\textbf{1. Model specification}

Let's consider the following random walk plus noise model:
\begin{eqnarray*} 
& Y_t = \theta_t + v_t \quad            & v_t \overset{i.i.d.}\sim N(0, 2000) \\
& \theta_t = \theta_{t-1} + w_t \quad   & v_t \overset{i.i.d.}\sim N(0, 8000)
\end{eqnarray*}
where, differently from the previous exercise, we set the process variances equal to $V=\sigma^2_v=2000$ and $W=\sigma^2_w=8000$.
The \textit{signal-to-noise ratio} is thus $r=\frac{W}{V}=\frac{8000}{2000}=4$, whereas with the original data it was only $r=\frac{W}{V}=\frac{1470}{15100}=0.097$.

The prior distribution is the same as before, that is:
\begin{align*} 
\theta_0 \sim N(1000, 1000)
\end{align*}

\textbf{2. Filtering}

Given this model specification, we compute the filtering state estimates $m_t =\mathbb{E}\;(\theta_t \mid y_{1:t})$ for $t=1,2,...,T$ obtained by implementing the Kalman filter. The estimates $(m_1, \ldots, m_T)$ are plotted in Figure \ref{fig:filter_est_2}.
\bigskip

```{r}
# Polinomial specification of the RW + noise
rw_Nile_2 <- dlmModPoly(order=1,dV=2000, dW=8000, m0=1000, C0=1000)

# Kalman filter
filt_Nile_2 <- dlmFilter(Nile,rw_Nile_2)

# Remove the first item (m_0) in the vector of filtered states
filt_est_2=dropFirst(filt_Nile_2$m)

# Compute the vector of state variances from the Singular Variance Decomposition matrices
list_c_2 <- dlmSvd2var(filt_Nile_2$U.C, filt_Nile_2$D.C)

# Compute standard deviations
vol_C_2 <- sqrt(unlist(list_c_2))
```

```{r, fig.width=5, fig.height=2, fig.cap="\\label{fig:filter_est_2} Filtering states estimates"}
# Df
df_Nile_filt_2 <-data.frame(years_Nile, Nile, filt_est_2)

# Plot
ggplot(df_Nile_filt_2, aes(x=years_Nile)) +
  geom_line(aes(y=filt_est_2), color = "red", size=0.4) +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y="State level") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(500,1500,300), limits = c(400,1500)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

The standard deviations $\sqrt{C_{t}} = V(\theta_t \mid y_{1:t-1})^{\frac{1}{2}}$ obtained using the Singular Value Decomposition procedure are reported in Figure \ref{fig:filter_sd_2}.
\bigskip

```{r}
# Sd
vol_C_plot_2 <- vol_C_2[2:101]
```

```{r, fig.width=5, fig.height=2, fig.cap="\\label{fig:filter_sd_2} Standard deviations of states estimates"}
# Df
df_Nile_sd_2 <-data.frame(years_Nile, vol_C_plot_2)

# Plot
ggplot(df_Nile_sd_2, aes(x=years_Nile)) +
  geom_line(aes(y=vol_C_plot_2), color = "darkgreen", size=0.4) +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y="Standard deviation") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

In Figure \ref{fig:filter_est_ci_2} we plot the 95% level Bayesian credible intervals for the filtering states estimates.
\bigskip

```{r, fig.width=7, fig.height=2, fig.cap="\\label{fig:filter_est_ci_2} Filtering estimates with credible intervals"}

upper_2 <- filt_est_2 + 1.96*vol_C_plot_2 
lower_2 <- filt_est_2 - 1.96*vol_C_plot_2 

df_Nile_ci_2 <-data.frame(years_Nile, Nile, filt_est_2, upper_2, lower_2)

colors <- c("Observed data" = "black", "Filtering estimates with C.I." = "red")

 ggplot(df_Nile_sd_2, aes(x=years_Nile))+
  geom_line(aes(y=Nile, color = "Observed data"), size=0.3) +
  geom_line(aes(y=filt_est_2, color = "Filtering estimates with C.I."), size=0.5) +
  geom_line(aes(y=upper_2), color = "red", linetype = "dashed", size=0.1) +
  geom_line(aes(y=lower_2), color = "red", linetype = "dashed", size=0.1) +
  geom_ribbon(aes(ymin=lower_2, ymax=upper_2), alpha=.1, fill="red") +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y=expression('Annual flow (10'^"8"*' m'^"3"*')'), color = "") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(400,1600,400), limits = c(250,1700)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))+
      theme(legend.justification = c("left", "top"),
              legend.box.just = "left",legend.margin = margin(2, 2, 2, 2))
```
\bigskip

\textbf{3. Forecasting}

Finally, in Figure \ref{fig:forecast_ci_2} we plot Nile time series with the one-step-ahead forecasts, i.e. $\hat{y}_{t}=\mathbb{E}\;({Y_{t} \mid y_{1:t-1})}=f_t$, and the credible intervals at the 95% level.
\bigskip

```{r}
# Create the vector of means for the observation process.
# Notice that here we do not have to remove the first element of the series
mean_forecast_2 <- filt_Nile_2$f

# Obtain the variance Q
filt_Q_2 <- unlist(list_c_2) + as.vector(rw_Nile_2$V) + as.vector(rw_Nile_2$W)
filt_Q_2 <- filt_Q_2[-1]

# Create upper and lower bounds of the credible interval
upper_for_2 <- mean_forecast_2 + 1.96*sqrt(filt_Q_2) 
lower_for_2 <- mean_forecast_2 - 1.96*sqrt(filt_Q_2) 
```

```{r, fig.width=7, fig.height=2, fig.cap="\\label{fig:forecast_ci_2} One-step ahead forecasts with credible intervals"}
# Df
df_nile_for_2 <-data.frame(years_Nile, Nile, mean_forecast_2, upper_for_2, lower_for_2)

colors <- c("Observed data" = "black", "Forecast with C.I." = "blue")

# Plot
ggplot(df_nile_for_2, aes(x=years_Nile)) +
  geom_line(aes(y=Nile, color = "Observed data"), size=0.3) +
  geom_line(aes(y=mean_forecast_2, color = "Forecast with C.I."), size=0.5) +
  geom_line(aes(y=upper_for_2), color = "blue", linetype = "dashed", size=0.1) +
  geom_line(aes(y=lower_for_2), color = "blue", linetype = "dashed", size=0.1) +
  geom_ribbon(aes(ymin=lower_for_2, ymax=upper_for_2), alpha=.1, fill="blue") +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y=expression('Annual flow (10'^"8"*' m'^"3"*')'), color = "") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(400,1600,400), limits = c(250,1700)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))+
      theme(legend.justification = c("left", "top"),
              legend.box.just = "left",legend.margin = margin(2, 2, 2, 2))
```

As we could expect, a low signal-to-noise ratio (as of Figures \ref{fig:filter_est_ci} and \ref{fig:forecast_ci}) implies a rather smooth state process (the signal) against rather imprecise observation process (the noise). On the contrary, a high signal-to-noise ratio (as in Figures \ref{fig:filter_est_ci_2} and \ref{fig:forecast_ci_2}) implies a noisy state process against a rather precise observation process.
Moreover, comparing figures from Questions 1 and 2 with those from Question 3, we notice that credible intervals have been affected as well by the change in the signal-to-noise ratio. That is not only due to $m_t$ (and consequently $f_t$) being affected by variances'change through the adaptive coefficient, as just explained, but also due to the the variances $C_t$ and $Q_t$.

Finally, comparing Figures \ref{fig:filter_sd} and \ref{fig:filter_sd_2}, we notice once again that the filtered standard deviation, $\sqrt{C_t}$ in the latter converges at a lower limit level than in the former (doing so also with a higher speed of convergence).
