---
title: "20236 Time Series Analysis - Assignment 6"
author:
- Simone Arrigoni (1794692)
- Luca Badolato (3086040)
- Simone Valle (3088281)
subtitle: "Bocconi University"
date: May 4, 2020
output: pdf_document

header-includes:
  \usepackage[utf8]{inputenc}
  \usepackage{setspace}
  \usepackage{apacite}
  \usepackage{natbib}
  \usepackage{algpseudocode}
  \usepackage{algorithm}
  \usepackage{bm}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{graphicx}
  \usepackage{subfig}
  \usepackage{booktabs, caption}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{listings}
  \usepackage{physics}
  \usepackage{float}
  
  \floatplacement{figure}{H}
  \usepackage{color} %red, green, blue, yellow, cyan, magenta, black, white
  \definecolor{mygreen}{RGB}{28,172,0} % color values Red, Green, Blue
  \definecolor{mylilas}{RGB}{170,55,241}
  \definecolor{codegray}{RGB}{0.5,0.5,0.5}
  \definecolor{orange}{RGB}{255,165,0}
  \DeclareMathOperator*{\E}{\mathbb{E}}
  \DeclareMathOperator*{\Ec}{\mathbb{E}_t}
  \setlength\parindent{0pt}
  
---

```{r, include=FALSE}

# Load useful packages
library(utf8)
library(labeling)
library(rmarkdown)
library(httr)
library(knitr)
library(tseries)
library(tinytex)
library(scales)
library(dlm)
library(magrittr)
library(stringr)
library(depmixS4)
library(tidyverse)
library(ggthemes)
library(latex2exp)
library(kableExtra)
library(ggpubr)
library(reshape2)
library(dlm)
# Settings
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      echo    = FALSE,
                      include = TRUE,
                      fig.pos = "H",
                      fig.align = "center",
                      out.width='60%')
```

\section{Exercise}

In this exercise we consider the \textbf{Nile} time series, available in R, which contains the measurements of the annual flow of the river Nile at Ashwan, for the period 1871-1970.

We first plot the data in Figure \ref{fig:Nile_plot}.
\bigskip

```{r, fig.width=5, fig.height=2, fig.cap="\\label{fig:Nile_plot}Annual flow of the river Nile at Ashwan"}
# Df
years_Nile<-seq(as.Date("1871/01/01"),as.Date("1970/01/01"), "years")
df_Nile<-data.frame(years_Nile,Nile)

# Plot
ggplot(df_Nile, aes(x=years_Nile, y=Nile)) +
  geom_line(color="black", size=0.4) +
  geom_vline(xintercept=as.Date('1899/01/01'), color="black", size=0.3)+
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y=expression('Annual flow (10'^"8"*' m'^"3"*')')) +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(500,1500,300), limits = c(400,1500)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))
```

The series is clearly non-stationary and presents a change point around 1998 (black vertical line). For this reason, studying it fitting a local level model seems a good idea.

\subsection{Question 1: Model definition and Smoothing}

Let's model the series with a random walk plus noise (or local level) model. This is a time-invariant DLM with univariate Gaussian observation and state process, i.e. $\underset{1 \times 1}{Y_t}$ and $\underset{1 \times 1}{\theta_t}$, as well as with constant system matrices $\underset{1 \times 1}F=1$ and $\underset{1 \times 1}G=1$:
\begin{eqnarray*} 
& Y_t = \theta_t + v_t \quad            & v_t \overset{i.i.d.}\sim N(0, V) \\
& \theta_t = \theta_{t-1} + w_t \quad   & v_t \overset{i.i.d.}\sim N(0, W)
\end{eqnarray*}
where the variances of both processes are constant and the prior distribution is also assumed to be Gaussian with mean $m_0$ and variance $C_0$. Furthermore, we assume that the error sequences $(v_t)$ and $(w_t)$ are independent, both within them and between them. We estimate the unknown variances $V=\sigma_v^2$ and $W=\sigma_w^2$ by maximum likelihood setting, with a slightly abuse of freedom, the mean of the initial distribution $m_0$ equal to the first observation.

```{r} 
#definition of the funtion of the rw which for each possible parameters build the dlm of interest. We assume that our initial distibution is centered at the first actual observation for simplicity. 
buildrw <- function(param){
    dlmModPoly(order=1, dV=param[1], dW=param[2], m0=Nile[1])}

#MLE computation
outMLE <- dlmMLE(Nile, parm = rep(100, 2), buildrw, lower=c(0.00001, 0), hessian=TRUE)

#output of the optimization algorithm:
#names(outMLE)
#outMLE$par  # MLE estimates 
#outMLE$value  # negative loglikelihood $-loglik(MLE)$
#outMLE$counts # number of iterations of the optimization algorithm
#outMLE$convergence # an integer code. 0 indicates successful convergence
#outMLE$hessian #Hessian matrix of - liglik(MLE)
#AsymCov=solve(outMLE$hessian) #asymptotic covariance matrix of the MLEs
#sqrt(diag(AsymCov)) # asymptotic standard errors

#model
NileMod <- buildrw(outMLE$par)
#NileMod$V
#NileMod$W
```

It is possible to estimate the unknown variances through the package *dlm*. Without entering into details, we create a function (**buildrw**) that for each possible unknown parameters (the variances $V=\sigma_v^2$ and $W=\sigma_w^2$  in our case) computes the likelihood, which is maximized using numerical optimiziation (via the R function **optim**). Furthermore, the asymptotic standard erros of the MLEs are based on the Hessian matrix of the loglikelihood, computed at the MLEs. We obtained the following estimates (standard errors in parentheses):

\begin{eqnarray*} 
\hat{v_t} = \hat{V}= \hat{\sigma^2_v} = 15098.7 \quad (4076.131) \\
\hat{w_t} = \hat{w} = \hat{\sigma^2_w} =  1469.021 \quad (1398.101) .
\end{eqnarray*} 

Having estimated the variances, we can now compute and plot the smoothing state estimates $s_t =\mathbb{E}\;(\theta_t \mid y_{1:T})$ obtained by implementing the Kalman smoother. Indeed, we want to retrospectively reconstruct the behaviour of the system given the full (observed) series up to time $T$. Using the Kalman smoother we can use a backward-recursive algorithm to compute the conditional distribution $\pi(\theta_t \mid y_{1:T})$) for any $t < T$, starting from the filtering distribution $\pi(\theta_T \mid y_{1:T})$) and estimating backward the states' history. We plot in figure \ref{fig:smoothing} the smoothing state estimates with $90 \%$ credible intervals. 


```{r, fig.width=9, fig.height=2.5, fig.cap="\\label{fig:smoothing} Smoothing states estimates with credible intervals"}
# Df
years_Nile<-seq(as.Date("1871/01/01"),as.Date("1970/01/01"), "years")

#Kalman Filter
NileFilt<-dlmFilter(Nile,NileMod)
m_T<-NileFilt$m[length(as.list(NileFilt$m))]
filt_est=dropFirst(NileFilt$m)

#Kalman Smoother
NileSmooth<-dlmSmooth(NileFilt)
s_T<-NileSmooth$s[length(as.list(NileSmooth$s))]
smooth_est<-NileSmooth$s[-1]

# Obtain the Smoothed variances
S<-dlmSvd2var(NileSmooth$U.S,NileSmooth$D.S)
sqrtS<-sqrt(unlist(S))
sqrtS_est<-sqrtS[-1]

# Calculate and plot confidence bands at 90% level
upper <- smooth_est + 1.65*sqrtS_est 
lower <- smooth_est - 1.65*sqrtS_est

df_Nile_ci <-data.frame(years_Nile, Nile, smooth_est, upper, lower)
   
colors <- c("Observed data" = "black", "Smoothed estimates" = "blue")

 ggplot(df_Nile_ci, aes(x=years_Nile))+
  geom_line(aes(y=Nile, color = "Observed data"), size=0.3) +
  geom_line(aes(y=smooth_est, color = "Smoothed estimates"), size=0.5) +
  geom_line(aes(y=upper), color = "red", linetype = "dashed", size=0.01) +
  geom_line(aes(y=lower), color = "red", linetype = "dashed", size=0.01) +
  geom_ribbon(aes(ymin=lower, ymax=upper), alpha=.1, fill="red") +
  geom_vline(xintercept = as.Date('1899/01/01'), color= "black", size=0.3) +
  theme_classic(base_size=9.5) +
  theme(panel.grid.minor = element_line(size=0.5)) +
  labs(x="Year", y=expression('Annual flow (10'^"8"*' m'^"3"*')'), color = "") +
  scale_x_date(date_breaks="20 years", limits=as.Date(c('1871/01/01','1970/01/01')),
               labels=date_format("%Y")) +
  scale_y_continuous(breaks = seq(500,1500,300), limits = c(400,1500)) +
  theme(axis.title.x=element_text(family="Times", margin=margin(t=5)),
        axis.title.y=element_text(family="Times", margin=margin(r=5)))+
   scale_color_manual(values = colors)+
        theme(legend.justification = c("left", "top"),
              legend.box.just = "left",legend.margin = margin(2, 2, 2, 2))


```

\subsection{Question 2: Model checking}