---
title: "20236 Time Series Analysis - Final project"
author:
- Simone Arrigoni (1794692)
- Luca Badolato (3086040)
- Simone Valle (3088281)
subtitle: "Bocconi University"
date: "21/4/2020"
output: pdf_document
geometry: margin=2cm
nocite: '@*'
bibliography: References.bib
header-includes:
  \usepackage[utf8]{inputenc}
  \usepackage{setspace}
  \usepackage{apacite}
  \usepackage{natbib}
  \usepackage{algpseudocode}
  \usepackage{algorithm}
  \usepackage{bm}
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{graphicx}
  \usepackage{subfig}
  \usepackage{booktabs, caption}
  \usepackage{array}
  \usepackage{threeparttable}
  \usepackage{listings}
  \usepackage{physics}
  \usepackage{float}
  
  \floatplacement{figure}{H}
  \usepackage{color} %red, green, blue, yellow, cyan, magenta, black, white
  \definecolor{mygreen}{RGB}{28,172,0} % color values Red, Green, Blue
  \definecolor{mylilas}{RGB}{170,55,241}
  \definecolor{codegray}{RGB}{0.5,0.5,0.5}
  \definecolor{orange}{RGB}{255,165,0}
  \DeclareMathOperator*{\E}{\mathbb{E}}
  \DeclareMathOperator*{\Ec}{\mathbb{E}_t}
  \setlength\parindent{0pt}
  
---

```{r, include=FALSE}

# Load useful packages
library(utf8)
library(labeling)
library(rmarkdown)
library(httr)
library(knitr)
library(tseries)
library(tinytex)
library(scales)
library(dlm)
library(magrittr)
library(stringr)
library(depmixS4)
library(tidyverse)
library(ggthemes)
library(latex2exp)
library(kableExtra)
library(ggpubr)
library(reshape2)
library(dlm)
# Settings
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      echo    = FALSE,
                      include = TRUE,
                      fig.pos = "H",
                      fig.align = "center",
                      out.width='80%')
```

\section{Introduction}

Pollution is one of the most debated social issues of our times. It has been attracting more and more attention given its serious implications on the health of both the Earth and human beings. The economic activity under the model of production and consumption that has been characterizing the last century is, with no doubt, at the very basis of such complex phenomenon. Only in the last fifteen years some consciousness about the importance of pollution in the long-run has evolved and serious interventions have been made. Moreover, among the numerous component which pollution can be split into, air pollution is certainly one of the most important and tightly linked to strict economic activity.

Therefore, the purpose of the present work is to model through common statistical techniques the evolution of air pollution during this last crucial fifteen years in a developped economy, i.e. the United States, trying to grasp some intuition about this fundamental phenomenon.

\section{Data}

The dataset \texttt{US\_pollution} contains monthly and daily data about air pollution for each state in the United States. It is provided by the United States Environmental Protection Agency (EPA). Specifically, four key indicators of air pollution are considered: 

\begin{itemize}
\item \textbf{Ground level ozone $\text{O}_{3}$} (\texttt{O3}): Bad ozone forms near the ground when pollutants (emitted by sources such as cars, power plants, industrial boilers, refineries, and chemical plants) react chemically in sunlight. Since it is more likely to form during warmer months, we expect to see a seasonal component. 
\item \textbf{Nitrogen dioxide $\text{NO}_{2}$} (\texttt{NO2}): It is produced as a result of road traffic and other fossil fuel combustion processes. Furthermore, its presence in air contributes to the formation and modification of other air pollutants, such as particle pollution.
\item \textbf{Carbon monoxide $\text{CO}$} (\texttt{CO}):  It forms from incomplete combustions mainly from vehicle exhausts (roughly 75\% of all carbon monoxide emissions nationwide and up to 95\% in cities), fuel combustion in industrial processes and natural sources. Cold temperatures make combustion less complete, therefore we expect also in this case a seasonal behavior. 
\item \textbf{Sulfur dioxide $\text{SO}_{2}$} (\texttt{SO2}): It is produced when sulfur-containing fuels are burned. It is common mainly in large industrial complexes. 
\end{itemize}
\medskip

Hence, the four variables of interest are interconnected and are the building blocks of a more general phenomenon, that is air pollution. Furthermore, they are related by a common latent process that can be summarized by several components such as industrialization, car traffic and regulation.

The preliminary analysis that led us from the raw data (with frequency of 4-intra-day observations) to the current dataset is made available in the GitHub repository at https://github.com/SimoneArrigoni/US_pollution. We relied on a moving average algorithm of order $120$ to smooth the data, assuming months of 30 days and a year of 360 days, with a minor approximation relative to the original data structure based on a actual 365 day-counts.

```{r echo=FALSE}
# Import the dataset
louisiana_df <- read.csv(url("https://raw.githubusercontent.com/SimoneArrigoni/US_pollution/master//Louisiana_dataset_monthly.csv"), sep = ",", header=T)
```

We will be focusing on measures for \textbf{Louisiana}, recorded monthly from 01/01/2000 to 27/04/2016, for a total of `r length(louisiana_df$n_months)`  observations. The main advantage of studying the evolution of air pollution in the same country over the years is that this allows us to control for fixed effects and to better study the latent process of interest that relates the four pollution components. The first five observations of the dataset are reported in Table \ref{tab:louisiana_df}.

```{r}
# create a vector of sates
n_dates<-seq(as.Date("2000/1/1"), as.Date("2016/6/1"), "month")
# How the dataset looks like
knitr::kable(louisiana_df[1:5, 1:5], "latex", booktabs=T,
             caption="Dataset structure\\label{tab:louisiana_df}") %>%
  kable_styling(latex_options="hold_position")
```
\bigskip

We report in Table \ref{tab:louisiana_corr} pollutants'correlation coefficients.

Provided that the correlation between pollutants show significant differences in regional patterns and that $\text{CO}$ is a good indicator for industrial and biomass burning pollution [@Logan1981], a positive and strong correlation with $\text{O}_{3}$ indicates that a region has experienced photochemical $\text{O}_{3}$ production from its precursors (including $\text{CO}$) and is aligned with previous studies in the field [@Voulgarakis2011]. $\text{CO}$ is also positively correlated with $\text{SO2}$ which is justified by the fact that they share a main source of production, i.e. coal extraction and combustion.

Other studies report that, in the US, $\text{O}_3$ is on average positively correlated with $\text{NO}_2$, which is opposite to our empirical findings about Louisiana, where they are slightly negatively correlated. A possible explanation can be given by substitution effects: an increase of nitrogen dioxide levels can be associated to the use of more fossil fuels that produce $\text{NO}_2$ instead of others. This might be related to Louisiana's fixed effects, however further rasearches would be beyond the purpose of our study.

```{r}
# Correlation plot
datacor = subset(louisiana_df, select=c(m_O3_mean, m_NO2_mean, m_CO_mean, m_SO2_mean))
cor.datacor = cor(datacor, use="complete.obs")
cormat = round(cor(datacor),2)
row.names(cormat) <- c("$\\text{O}_{3}$",
                       "$\\text{NO}_{2}$",
                       "$\\text{CO}$",
                       "$\\text{SO}_{2}$")

knitr::kable(cormat, "latex", booktabs=T, align="c", escape=F,
             caption="Correlation table \\label{tab:louisiana_corr}",
             col.names=c("$\\text{O}_{3}$",
                         "$\\text{NO}_{2}$",
                         "$\\text{CO}$",
                         "$\\text{SO}_{2}$")) %>%
  column_spec(1:4, width="5em") %>%
kable_styling(latex_options="hold_position")
```


We plot in Figure \ref{fig:louisiana_plots} the time series for the four pollutants under study to give a first qualitative description of the phenomenon.

```{r, fig.width=20, fig.height=6, out.width='100%', fig.cap="\\label{fig:louisiana_plots} Air pollution in the state of Louisiana"}

# Time series plot
O3 <- ggplot(louisiana_df, aes(x=n_months, y=m_O3_mean)) +
  geom_line(color="steelblue", size=0.7) +
  theme_classic(base_size=9) +
  labs(y=expression(O[3]~(ppm)), x="Year") +
  scale_x_continuous(breaks=seq(0,200,50),labels=format(seq(as.Date("2000/1/1"),as.Date("2016/6/1"), "4 years"), "%Y")) +
  theme(axis.text=element_text(size=19),
        axis.title.x=element_text(size=21,family="Times", margin=margin(t=5)),
        axis.title.y=element_text(size=21,family="Times", margin=margin(r=5)))

NO2 <- ggplot(louisiana_df, aes(x=n_months, y=m_NO2_mean)) +
  geom_line(color="steelblue", size=0.7) +
  theme_classic(base_size=9) +
  labs(y=expression(NO[2]~(ppb)), x="Year") +
  scale_x_continuous(breaks=seq(0,200,50),labels=format(seq(as.Date("2000/1/1"),as.Date("2016/6/1"), "4 years"), "%Y")) +
  theme(axis.text=element_text(size=19),
        axis.title.x=element_text(size=21,family="Times", margin=margin(t=5)),
        axis.title.y=element_text(size=21,family="Times", margin=margin(r=5)))

CO <- ggplot(louisiana_df, aes(x=n_months, y=m_CO_mean)) +
  geom_line(color="steelblue", size=0.7) +
  theme_classic(base_size=9) +
  labs(y=expression(CO~(ppm)), x="Year") +
  scale_x_continuous(breaks=seq(0,200,50),labels=format(seq(as.Date("2000/1/1"),as.Date("2016/6/1"), "4 years"), "%Y")) +
  theme(axis.text=element_text(size=19),
        axis.title.x=element_text(size=21,family="Times", margin=margin(t=5)),
        axis.title.y=element_text(size=21,family="Times", margin=margin(r=5)))

SO2 <- ggplot(louisiana_df, aes(x=n_months, y=m_SO2_mean)) +
  geom_line(color="steelblue", size=0.7) +
  theme_classic(base_size=9) +
  labs(y=expression(SO[2]~(ppb)), x="Year") +
  scale_x_continuous(breaks=seq(0,200,50),labels=format(seq(as.Date("2000/1/1"),as.Date("2016/6/1"), "4 years"), "%Y")) +
  theme(axis.text=element_text(size=19),
        axis.title.x=element_text(size=21,family="Times", margin=margin(t=5)),
        axis.title.y=element_text(size=21,family="Times", margin=margin(r=5)))

ggarrange(O3, NO2, CO, SO2,
          ncol=2, nrow=2)
```
The time series, each with its specific features, do not in general look stationary.  Hence, they cannot be analized with the common models based on the assumption of covariance-stationarity, such as the well-known ARMA models.

\bigskip

\section{$\text{CO}$ series}
$\text{CO}$ is perhaps the pollutant more tightly linked with the economic and industrial network of a country. Consequently, we select this as the series of main interest for our analysis, being the purpose of this work tracking the evolution of air pollution relative to an underlying economic process.

As exploratory step, we decompose the time series in its structural component by using an additive model. Figure \ref{fig:decompose_CO}, that plots together the time series with such components, shows that, in fact, a seasonal factor seems to be present. Secondly, the series is characterized by a negative trend up to a clear breakponit in september 2007 (slightly before month 100). The effect of this break seems to be twofold: firstly, the trend, that was negative in the pre-break period, disappears; secondly, the variance reduces markedly in the post-break period.
These features suggest that the process may be described by a Homogeneous Hidden Markov Model with a trend component.  

```{r fig.width=18, fig.height=8, out.width='60%', fig.cap="\\label{fig:decompose_CO}Additive decomposition into structural components"}

ts_CO<-ts(louisiana_df$m_CO_mean,start=c(2000,1,1),end = c(2016,6,1),frequency = 12)
dec_CO<-decompose(ts_CO,type=c("additive"))
# plot(dec_CO,cex.main=2.5, cex.lab=1.6, cex.axis=2)

# Convert dec_CO to data frame
dat = cbind(n_dates, with(dec_CO, data.frame(Obs=x, Trend=trend, Seas=seasonal, Rand=random)))

ggplot(gather(dat, component, value, -n_dates), aes(n_dates, value)) +
  facet_grid(component ~ ., scales="free_y") +
  geom_line(size=0.7) +
  theme_bw() +
  labs(y=expression(CO[2]~(ppm)), x="Year") +
  ggtitle(expression(Decomposed~CO[2]~Time~Series)) +
  theme(plot.title=element_text(size=25,family="Times",face="bold",hjust=0.5),
        axis.text=element_text(size=21),
        axis.title.x=element_text(size=23,family="Times", margin=margin(t=5)),
        axis.title.y=element_text(size=23,family="Times", margin=margin(r=5)),
        strip.text=element_text(size=21,family="Times",face="bold"))
```

\subsection{HMM for modelling trend}
As outlined in the previous section, given the features of the $\text{CO}$ series, a suitable HMM with two states may be specified as follows: 

\begin{align*}
\begin{cases}
Y_{t}=\mu_{1} + \beta_{1}t + \varepsilon_{t}, \quad \varepsilon_{t}\overset{i.i.d.}{\sim}N(0,\sigma_{1}^2) \qquad \text{if} \; S_{t}=1 \\
Y_{t}=\mu_{2}+ + \beta_{2}t + \varepsilon_{t}, \quad \varepsilon_{t}\overset{i.i.d.}{\sim}N(0,\sigma_{2}^2) \qquad \text{if} \; S_{t}=2 \\
\end{cases}
\end{align*}

where there are three state-dependent parameters, namely the level, $\mu_i$, the slope, $\beta_i$, and the variance, $\sigma_i$, for $i=1,2$. If our guess was correct, we would expect $\mu_1 > \mu_2$, $\beta_1<0$, $\beta_2=0$ and $\sigma_1^2>\sigma_2^2$ .

```{r echo=FALSE}
# Model specification
y <- as.numeric(louisiana_df$m_CO_mean)
nstates <- 2
set.seed(2)
HMM_CO <- depmixS4::depmix(y ~ 1 + louisiana_df$n_months, data=data.frame(y), nstates=nstates)
```

```{r include=FALSE}
# Estimation of the unknown parameters
HMM_CO_fit <- depmixS4::fit(HMM_CO)
```

```{r}
# MLE
mu1hat <- paste0(round(HMM_CO_fit@response[[1]][[1]]@parameters[["coefficients"]][["(Intercept)"]],3))
mu2hat <- paste0(round(HMM_CO_fit@response[[2]][[1]]@parameters[["coefficients"]][["(Intercept)"]],3))

beta1hat<-paste0(round(HMM_CO_fit@response[[1]][[1]]@parameters[["coefficients"]][["louisiana_df$n_months"]],3))
beta2hat<-paste0(round(HMM_CO_fit@response[[2]][[1]]@parameters[["coefficients"]][["louisiana_df$n_months"]],3))

sd1hat <- paste0(round(HMM_CO_fit@response[[1]][[1]]@parameters$sd,3))
sd2hat <- paste0(round(HMM_CO_fit@response[[2]][[1]]@parameters$sd,3))

# se(MLE)
MLE_st.err=depmixS4::standardError(HMM_CO_fit)
st.err_mu1hat <- paste0('(', round(MLE_st.err$se[7],3), ')')
st.err_mu2hat <- paste0('(', round(MLE_st.err$se[10],3), ')')

st.err_beta1hat<-paste0('(', round(MLE_st.err$se[8],3), ')')
st.err_beta2hat<-paste0('(', round(MLE_st.err$se[11],3), ')')

st.err_sd1hat <- paste0('(', round(MLE_st.err$se[9],3), ')')
st.err_sd2hat <- paste0('(', round(MLE_st.err$se[12],3), ')')

```

```{r}
# Build a summary table for estimates
MLEsum <- data.frame(state=c(rep("S=1",2), rep("S=2",2)),
                     mu=c(mu1hat, st.err_mu1hat, mu2hat, st.err_mu2hat),
                     beta=c(beta1hat, st.err_beta1hat,beta2hat,st.err_beta2hat),
                     sd=c(sd1hat, st.err_sd1hat, sd2hat, st.err_sd2hat))
knitr::kable(MLEsum, "latex", booktabs=T, align="c",
             col.names = c("State", "$\\hat{\\mu}_{MLE}$","$\\hat{\\beta}_{MLE}$","$\\hat{\\sigma}_{MLE}$"),
             escape=F, caption="Maximum Likelihood estimates of the state-dependent
             parameters \\label{tab:HMM_MLE}") %>%
  column_spec(1:4, width="5em") %>%
  collapse_rows(columns=1, latex_hline="none") %>%
  footnote(general="Standard errors in parentheses",
           general_title="Note:",
           footnote_as_chunk=T, escape=F) %>%
  kable_styling(latex_options="hold_position")

```

Table \ref{tab:HMM_MLE} reports, however, unexpected results. Indeed, the slope estimate in the first state, $\hat{\beta_1}_{MLE}$, is only slightly negative and very close to zero, whereas the estimate for the second state, $\hat{\beta_2}_{MLE}$ is null as we expected. Overall, the modell suggest that the trend in the two states is not different (i.e. it is equal to zero).
The change point, instead, significantly affects both the level and the variance. In fact, what seemed to be a change in the trend is, instead, a drop in the level, i.e. $\hat{\mu_1}_{MLE}>\hat{\mu_2}_{MLE}$, followed by a sharp decline in variance,i.e. $\hat{\sigma_1}_{MLE}>\hat{\sigma_2}_{MLE}$.

```{r}
# Decoding
# Get the estimated state for each timestep 
estStates_CO <- posterior(HMM_CO_fit)
```

The series with HMM estimated state-dependent means and state-dependent standard deviations are shown in Figure \ref{fig:HMM_results} (left panel) along with the Viterbi states obtained through \textit{global decoding} and with the posterior probabilities (right panel).

```{r, include=FALSE}
HMM_summary<-depmixS4::summary(HMM_CO_fit)
```


```{r, fig.width=20, fig.height=6, out.width='90%', fig.cap="\\label{fig:HMM_results}State-dependent means and standard deviations"}

# Chart with state-dependent mean and standard deviations
HMM_main_CO <- data.frame(state=1:nstates,
                      mu_CO=HMM_summary[,1],
                      sigma_CO=HMM_summary[,3])
df_to_plot_CO <- estStates_CO %>%
  left_join(HMM_main_CO)
df_to_plot_CO %<>%
  mutate(xtime=louisiana_df$n_months, yvalue=louisiana_df$m_CO_mean)

HMM_main<- ggplot(df_to_plot_CO, aes(x=xtime, y=yvalue)) +
                geom_line(size=0.7) +
                geom_point(aes(x=xtime, y=mu_CO), col="blue", size=.8) +
                geom_ribbon(aes(ymin=mu_CO-2*sigma_CO, ymax=mu_CO+2*sigma_CO), alpha=.1) +
                theme_classic(base_size=9) +
                theme(plot.title=element_text(hjust=0.5)) +
                labs(y=expression(CO~(ppm)), x="Months") +
                scale_x_continuous(breaks=seq(0,200,50),labels=format(seq(as.Date("2000/1/1"),as.Date("2016/6/1"), "4 years"), "%Y")) +
                scale_y_continuous(breaks=seq(0,1.5,0.25)) +
                theme(plot.title=element_text(size=10, face="bold", margin=margin(0,0,10,0)),
                axis.text=element_text(size=19,face="bold"),
                axis.title.x=element_text(size=21,family="Times", margin=margin(t=5)),
                axis.title.y=element_text(size=21,family="Times",margin=margin(r=5)))

# Chart with viterbi states and posterior probabilities
HMM_additional_CO <- data.frame(time_index=louisiana_df$n_months,
                           state=estStates_CO[1],
                           S1=estStates_CO[2]) %>% 
  gather("variable", "value", -time_index)

my_breaks <- function(x) { if (max(x)>1) seq(0,2,1) else seq(0,1,1) }
lab_names <- c('S1'="Posterior Prob.",'state'="Viterbi states")

HMM_additional<-ggplot(HMM_additional_CO, aes(time_index, value)) +
  geom_line(color="black", size=0.7) +
  facet_wrap(variable~., scales="free", ncol=1,
             strip.position="left",
             labeller = as_labeller(lab_names)) +
  labs(x="Month") +
  scale_y_continuous(breaks=my_breaks) +
  scale_x_continuous(breaks=seq(0,200,50), labels=format(seq(as.Date("2000/1/1"),as.Date("2016/4/27"), "4 years"), "%Y"))+
  theme_stata(base_size=9) +
  theme(axis.text=element_text(size=19),
        axis.title.x=element_text(size=21,family="Times", margin=margin(5,0,0,0)),
        axis.title.y=element_blank(),
        strip.text=element_text(size=21,family="Times",face="bold"),
        strip.background=element_rect(colour="black", fill=NA),
        strip.placement="outside")
# Option 1
# ggarrange(HMM_main, HMM_additional, ncol=1, widths = c(6,0.3), heights = c(1,0.8))

#option 2
ggarrange(HMM_main, HMM_additional,widths = c(2,1))
```

Explanations of these results, which confirm a stable change in the evolution process of CO in Louisiana, are several. First, the decline of carbon particles has been widely attributed to a shift from the use of coal to natural gas in US electricity production which dates back exactly to 2007. This is also supported by a marked decline in US coal mines activity which started in that same year. Secondly, a high and positive correlation between economic growth is well documented, thus highlighting some relationship with the outbreak of the Great Recession. Moreover, 2007 was a crucial year when negotiations on greenhouse gas emission were made at international level (such as at the G8 summit and the Vienn agreement). 

A deeper anlysis on the factors explaining this shift can be found in [@feng2015drivers], in which they analyse the serie of $\text{CO2}$.

\subsection{DLM - Univariate dynamic regression}

\textcolor{red}{DLM for Univariate regression potrebbe essere un'idea. per esempio come rgressore si potrebbe usare la quantità di carbone estratto negli USA, la quantità di gas naturale usata nella produzione di corrente. se trovare i dati fosse difficile potremmo semplicemente modellare il trend}

INSERT HERE EVERITHING ON MODEL 2 FOR CO

\section{DLM multivariate series}

\textcolor{red}{le serie SO2 e CO mostrano entrambe un breakpoint alla Month 400 e sono positivamente correlate. inoltre entrambi i pollutants dipendono dalla combustione del carbone che potremmo usare come regressore. in questo caso avremmo un multivariate DLM con m=2}

INSERT HERE MULTIVARIATE TIME SERIES MODEL

\section{Conclusion}

INSERT HERE CONCLUSIONS

\newpage
\section{References}